---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

## Required libraries
```{r libraries}
library(ggplot2)
library(lubridate)
```

## Loading and preprocessing the data
```{r dataload}
## csv file must be in working directory
rawdf <- read.csv('activity.csv', stringsAsFactors =F)
```

## What is mean total number of steps taken per day?
### Preparing data
```{r prep1}
### locate rows with NA values for steps
na_rows <- is.na(rawdf$steps)

### keep usable data
df <- rawdf[!na_rows,]

### convert string date field "date" to R date
df$day <- ymd(df$date)
```

### Histogram 
```{r plot1}
#### number of steps per day
steps_d <- aggregate(steps~day, data=df, sum)
g <- ggplot(data=steps_d, aes(steps_d$steps)) 
g <- g + geom_histogram(binwidth=500)
g + labs(title = "Histogram of the total number of steps per day"
         ,x="Steps per day", y="Count")
```

### Mean and median for number of steps per day
```{r summary1}
summary1<-summary(steps_d$steps)
summary1
```

## What is the average daily activity pattern?
### Time series plot of the average number of steps taken
```{r plot2}
#### aggregate steps on day and compute mean
steps_dm <- aggregate(steps~day, data=df, mean)
g <- ggplot(data=steps_dm, aes(x=day, y=steps)) 
g <- g + geom_line(size=1)
g + labs(title = "Time series plot of the average number of steps taken"
         ,x="day", y="Average count")
```

### 5 minute interval containing the maximum number of steps
```{r prep3}
### aggregate steps on interval and compute mean
steps_im <- aggregate(steps~interval, data=df, mean)

### index at which max value for steps is found
i_max <- match(max(steps_im$steps),steps_im$steps )
```

```{r max}
### interval containing the maximum number of steps
steps_im$interval[i_max]
```

## Imputing missing values
### total number of missing values in the dataset
```{r missing}
#### rows with NA values were identified earlier with na_rows is.na(rawdf$steps)
#### there are
length(na_rows)
#### missing values
```
### Strategy for imputing missing data
```{r prep4}
#### missing steps values for intervals are replaced by 
#### interval average values calculated earlier

#### subset for rows with na values
df_na <- rawdf[na_rows,]

#### join (merge) data frames df_na and steps_im calculated earlier
names(df_na)
names(steps_im)
df_joined <- merge(df_na[,c(2,3)],steps_im, by="interval" )
names(df_joined)
#### convert string date field "date" to R date
df_joined$day <- ymd(df_joined$date)

#### as names(df) order is different from names(df_joined)
cols<- intersect( names(df), names(df_joined) )

#### rows with average values are appended
df <- rbind(df[, cols], df_joined[,cols])
```

### Histogram of the total number of steps per day after missing values are imputed
```{r plot3}
#### aggregate steps on day
steps_d_filled <- aggregate(steps~day, data=df, sum)
g <- ggplot(data=steps_d_filled, aes(steps_d_filled$steps)) 
g <- g + geom_histogram(binwidth=500)
g + labs(title = "Histogram of the total number of steps per day, with imputed values"
         ,x="Steps per day", y="Average count")
```

### mean and median for number of steps per day including imputed valuues
```{r summary2}
summary2<-summary(steps_d_filled$steps)
summary2
```

### Do mean and median values differ from the estimates from the first part of the assignment?
```{r compar_mean_median}
#### without imputed values
summary1
#### with imputed values
summary2
#### mean remains the same, median differs
#### we notice an increase of 1 for the nedian
```

## Are there differences in activity patterns between weekdays and weekends?
### preparation
```{r prep5}
#### computes numeric day of week (Sunday is 1)
df$wday <- wday(df$day)

#### create a mapping day to type of day
day2dtype <- data.frame( wday=sort(unique(df$wday))
                        ,dtype=c( "WEEKEND","WEEKDAY","WEEKDAY","WEEKDAY"
                                ,"WEEKDAY","WEEKDAY","WEEKEND"))
day2dtype

#### join df with mapping
names(df)
names(day2dtype)
df_merged <- merge(df,day2dtype, by="wday")
names(df_merged)
```

### Time series plots of the average number of steps taken per interval
```{r plot4}
#### average number of steps per 5 minutes interval and day type
steps_idm <- aggregate(steps~interval+dtype, data=df_merged, mean)
g <- ggplot(steps_idm, aes(x=interval, y=steps)) 
g <- g + geom_line(size=1) + facet_wrap(~dtype, ncol=1)
g + labs(title = "Average number of steps taken per 5 minutes interval across week days and weekends"
         ,x="Interval", y="Average count")
```

### Activity patterns between weekdays and weekends differ
### On weekdays activity peaks in the morning
### On weekends activity spreads out more evenly during morning and afternoon.
